---
title: "Report: Design of guide RNAs with `snakemake-crispr-guides`"
auhtor: "`r sessionInfo()$platform`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
   rmd: "report.Rmd"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_depth: 2
    number_sections: yes
    df_print: paged
---

----------

# Background

This workflow is a best-practice workflow for the automated generation of guide RNAs for CRISPR applications. It's main purpose is to provide a simple, efficient and easy-to-use framework to design thousands of guides simultaneously for CRISPR libraries from as little input as an organism's name/genome ID. For the manual design of single guides, users are instead referred to even simpler web resources such as [Chop-Chop](http://chopchop.cbu.uib.no/), [CRISPick](https://portals.broadinstitute.org/gppx/crispick/public), or [Cas-OFFinder/Cas-Designer](http://www.rgenome.net/cas-designer/).

This workflow relies to a large degree on the underlying [Bioconductor package ecosystem `crisprVerse`](http://bioconductor.org/packages/release/bioc/html/crisprVerse.html), published in 2022 by:

> Hoberecht, L., Perampalam, P., Lun, A. et al. _A comprehensive Bioconductor ecosystem for the design of CRISPR guide RNAs across nucleases and technologies_. Nat Commun 13, 6568 (**2022**). https://doi.org/10.1038/s41467-022-34320-7.

The workflow is built using [snakemake](https://snakemake.readthedocs.io/en/stable/) and consists of the following steps:

1. Obtain genome database in `fasta` and `gff` format (`python`, [NCBI Datasets](https://www.ncbi.nlm.nih.gov/datasets/docs/v2/))
   1. Using automatic download from NCBI with a `RefSeq` ID
   2. Using user-supplied files
2. Find all possible guide RNAs for the given sequence, with many options for customization (`R`, `crisprVerse`)
3. Collect on-target and off-target scores (`R`, `crisprVerse`, `Bowtie2`)
4. Filter and rank guide RNAs based on scores and return final list (`R`, `crisprVerse`)
5. Generate report with overview figures and statistics (`R markdown`)

If you want to contribute, report issues, or suggest features, please get in touch on [github](https://github.com/MPUSP/snakemake-crispr-guides).

----------

# Prerequisites

## Packages

Loaded required R packages:

- `tidyverse`
- `ggrepel`
- `ggpubr`
- `GGally`


```{r, echo = FALSE, warning = FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggrepel)
  library(ggpubr)
  library(GGally)
})
```


```{r, echo = FALSE}
# custom ggplot2 theme that is reused for all later plots
custom_colors <- c("#E7298A", "#66A61E", "#E6AB02", "#7570B3", "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D")
custom_range <- function(n = 5) {
  colorRampPalette(custom_colors[c(1, 5, 2)])(n)
}

custom_theme <- function(base_size = 12, base_line_size = 1.0, base_rect_size = 1.0, ...) {
  theme_light(base_size = base_size, base_line_size = base_line_size, base_rect_size = base_rect_size) + theme(
    title = element_text(colour = grey(0.4), size = 10),
    plot.margin = unit(c(12, 12, 12, 12), "points"),
    axis.ticks.length = unit(0.2, "cm"),
    axis.ticks = element_line(colour = grey(0.4), linetype = "solid", lineend = "round"),
    axis.text.x = element_text(colour = grey(0.4), size = 10),
    axis.text.y = element_text(colour = grey(0.4), size = 10),
    panel.grid.major = element_line(linewidth = 0.6, linetype = "solid", colour = grey(0.9)),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(linetype = "solid", colour = grey(0.4), fill = NA, linewidth = 1.0),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(colour = grey(0.4), size = 10, margin = unit(rep(3, 4), "points")),
    legend.text = element_text(colour = grey(0.4), size = 10),
    legend.title = element_blank(),
    legend.background = element_blank(),
    ...
  )
}

# set graphical parameter for subfigure labels
list_fontpars <- list(face = "plain", size = 14)

# default figure size
figwidth <- 7.0
figheight <- 4.5

# function to export an image as svg and png
save_plot <- function(pl, path = "../figures/", width = 6, height = 6) {
  pl_name <- deparse(substitute(pl))
  svg(
    filename = paste0(path, pl_name, ".svg"),
    width = width, height = height
  )
  print(pl)
  dev.off()
  png(
    filename = paste0(path, pl_name, ".png"),
    width = width * 125, height = height * 125, res = 120
  )
  print(pl)
  invisible(capture.output(dev.off()))
}
```


## Result table

Imported table of predicted guide RNAs.

*NOTE: tables are only rendered in HTML output but not PDF.*

Head of the table (first six rows):

```{r, echo = FALSE}
df_guides <- read_csv(snakemake@input$guides, show_col_types = FALSE)
head(df_guides)
```

## Configuration

Show parameters that were used for guide RNA prediction.
The displayed parameters are taken from the pipelines `config.yml` file.

*NOTE: tables are only rendered in HTML output but not PDF.*

```{r, echo = FALSE}
# import some required parameters
tss_window <- snakemake@config$design_guides$tss_window
n_examples <- snakemake@config$visualize_guides$show_examples

# show all parameters for pipeline
list_config <- read_lines("config/config.yml")
list_config <- gsub("\"", "", list_config)
print(list_config)
```


----------

# Summary statistics

## Number of predicted small guide RNAs

- Overview about number of targeted genes or other elements
- Overview about number of predicted small guide RNAs (sgRNAs)
- Distribution of number of sgRNAs per target (gene)

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
df_n_guides <- df_guides %>%
  group_by(tx_name) %>%
  summarize(n_guides = n()) %>%
  count(n_guides)

print(paste0(
  "Total number of targets (genes) with predicted guide RNAs: ",
  length(unique(df_guides$tx_name))
))

print(paste0(
  "Total number of predicted guide RNAs: ",
  nrow(df_guides)
))

plot_n_guides <- df_n_guides %>%
  ggplot(aes(x = n_guides, y = n)) +
  geom_col(color = "white", fill = custom_colors[1]) +
  labs(
    x = "guides per target", y = "count",
    title = "Number of predicted guide RNAs per target"
  ) +
  custom_theme()

print(plot_n_guides)
```

## Strand, width, position, and PAM site

- Strand of guides is one of "+" (template strand) or "-" (non-template strand)


```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figwidth}
ggpubr::ggarrange(
  df_guides %>%
    count(strand) %>%
    ggplot(aes(x = factor(strand), y = n)) +
    geom_col(color = "white", fill = custom_colors[1]) +
    labs(x = "strand", y = "count", title = "Guides per strand") +
    custom_theme(),
  df_guides %>%
    count(width) %>%
    ggplot(aes(x = factor(width), y = n)) +
    geom_col(color = "white", fill = custom_colors[1]) +
    labs(x = "width", y = "count", title = "Width distribution of guides") +
    custom_theme(),
  df_guides %>%
    count(pam) %>%
    ggplot(aes(x = factor(pam), y = n)) +
    geom_col(color = "white", fill = custom_colors[1]) +
    labs(x = "pam", y = "count", title = "PAM sequence of guides") +
    custom_theme(),
  df_guides %>%
    ggplot(aes(x = dist_to_tss)) +
    geom_histogram(bins = 30, color = "white", fill = custom_colors[1]) +
    labs(x = "distance [nt]", y = "count", title = "Relative distance to TSS") +
    custom_theme()
)
```

## On-target scores

- What is the distribution of on-target scores?
- 'Good' distributions should be normally distributed between a range of 0 and 1
- 'Bad' distributions are too narrow, indicating that the score is probably not useful to differentiate between high and low efficacy

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
df_guides %>%
  select(matches("^score_") & !matches("^score_(cfd|mit)")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "score",
    values_to = "value",
    names_transform = function(x) gsub("score_", "", x)
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, color = "white", fill = custom_colors[1]) +
  labs(x = "score", y = "count", title = "Global distribution of on-target scores") +
  facet_wrap(~score) +
  custom_theme()
```

----------

- What is the correlation between scores?
- Depicted is the scatter plot of each score against each other, as well as the correlation coefficient _R_
- It is not unusual that scores show a low degree of correlation with each other
- However a slight positive correlation should be visible, a negative correlation would be problematic

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figwidth}
df_guides %>%
  dplyr::select(matches("^score_") & !matches("^score_(cfd|mit)")) %>%
  rename_with(.fn = function(x) gsub("score_", "", x)) %>%
  ggpairs(aes(color = "score")) +
  labs(
    x = "", y = "",
    title = "Correlation of scores with each other",
    subtitle = "Lower triangle: scatter plots, upper triangle: correlation coefficent R"
  ) +
  custom_theme()
```


## Location of guide RNAs

- The following figures show the position of guide RNAs on a selection of target genes
- The first figure shows the distribution of all realtive guide RNA positions
- The next figure shows a selection of the first 30 genes for the purpose of quality control

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
df_guides %>%
  dplyr::count(dist_to_tss) %>%
  ggplot(aes(x = dist_to_tss, y = n)) +
  geom_col(fill = custom_colors[1]) +
  geom_smooth(method = 'loess', formula = y ~ x,
    color = grey(0.3), fill = custom_colors[1]) +
  labs(
    x = "", y = "",
    title = "Distribution of all guide RNAs over the TSS window",
    subtitle = "Bars are sgRNAs counts per position, ribbon is smoothed average"
  ) +
  custom_theme()
```


```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = 10}
df_guides %>%
  filter(tx_name %in% unique(df_guides$tx_name)[1:n_examples]) %>%
  mutate(tx_name = paste0(tx_name, " (", tss_strand, ")") %>%
    factor(., unique(.))) %>%
  mutate(
    xstart = case_when(
      tss_strand == "+" & strand == "+" ~ start - tss_pos,
      tss_strand == "+" & strand == "-" ~ end - tss_pos,
      tss_strand == "-" & strand == "-" ~ -1 * (end - tss_pos),
      tss_strand == "-" & strand == "+" ~ -1 * (start - tss_pos),
    ),
    xend = case_when(
      tss_strand == "+" & strand == "+" ~ end - tss_pos,
      tss_strand == "+" & strand == "-" ~ start - tss_pos,
      tss_strand == "-" & strand == "-" ~ -1 * (start - tss_pos),
      tss_strand == "-" & strand == "+" ~ -1 * (end - tss_pos),
    )
  ) %>%
  mutate(y = ifelse(strand == tss_strand, 0.25, -0.25)) %>%
  ggplot(aes(x = xstart, y = y, xend = xend, yend = y)) +
  geom_segment(
    x = tss_window[1], y = 0, xend = 1.05 * tss_window[2], yend = 0,
    linewidth = 1.5, color = grey(0.7), arrow = arrow(
      angle = 30, length = unit(0.06, "inches"),
      type = "closed"
    )
  ) +
  geom_segment(aes(color = ifelse(strand == tss_strand, "+", "-")),
    linewidth = 1.5, arrow = arrow(
      angle = 30, length = unit(0.03, "inches"),
      type = "closed"
    )
  ) +
  facet_wrap(~tx_name, ncol = 1) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "distance to TSS [nt]", y = "",
    title = "Guide RNA location in TSS window, for selected genes",
    subtitle = "Arrows show direction of the guide. Note that genes on the '-' strand are shown as reverse-complement"
  ) +
  custom_theme(legend.position = 0) +
  scale_color_manual(values = custom_colors[1:2]) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

----------

# About this report

## Pipeline

This report was automatically generated by the [snakemake-crispr-guides](https://github.com/MPUSP/snakemake-crispr-guides) pipeline.

For issues, bugs, and feature requests please use the pipeline's [github page](https://github.com/MPUSP/snakemake-crispr-guides/issues).

For all other feedback, contact the author(s):

- Michael Jahn, PhD (author and maintainer): jahn@mpusp.mpg.de

The pipeline is developed at the [Max Planck Unit for the Science of Pathogens](https://www.mpusp.mpg.de/), Berlin, Germany.

## Packages used in the pipeline

This list shows the fixed dependencies used as `conda` environment variables within the pipeline.

```{r, echo = FALSE, warnings = FALSE}
versions <- read_delim("results/versions/log_packages.txt",
  delim = "\n",
  show_col_types = FALSE,
  col_names = "module"
)

print(versions, n = Inf)
```

## Data accessability

The following resources were used to generate this report:

- Guide RNA results table: `results/design_guides/guideRNAs_top.csv`
- config file: `config/config.yml`

## Session Info

Link to the R markdown source that was used to generate this report:

<a download="report.Rmd" href="`r base64enc::dataURI(file = params$rmd, mime = 'text/rmd', encoding = 'base64')`">R Markdown source file</a>

Session info (R base and package versions):

```{r, echo = FALSE}
sessionInfo()
```

```{r, echo = FALSE}
write_lines(
  "VISUALIZE_GUIDES: finished writing HTML report successfully",
  file = "results/visualize_guides/log.txt"
)
```